---
title: "Data Cleaning"
---
```{css, echo = FALSE}
.justify {
    text-align: justify !important;
    text-indent: 20px; 
}

.epigrafe {
    text-align: justify !important;
    text-indent: 20px; 
    border: 1.5px solid #87c8b5; 
    padding-top: 15px;
    padding-bottom: 5px;
    padding-right: 15px;
    padding-left: 15px;
    font-size: 14px;
    background-color: #f9f9f9; 
    margin: 20px 0px 30px 0px;
}
```

### Overview
::: {.justify}
In this section, we will perform the data cleaning process. In order to improve the quality and accuracy of each dataset collected in the Data Gathering section, we apply several steps.
:::

Below are the libraries used in this section:
``` {r}
#| echo: true
#| message: false
#| code-fold: true
#| code-summary: "R Libraries"
#| results: 'hide'
#| warning: false

library(tidyverse)
library(ggplot2)
library(forecast)
library(astsa)
library(xts)
library(tseries)
library(fpp2)
library(fma)
library(lubridate)
library(TSstudio)
library(quantmod)
library(tidyquant)
library(plotly)
library(gridExtra)
library(readxl)
library(imputeTS)
library(zoo)
library(knitr)
library(kableExtra)
library(patchwork)
library(vars)
library(dplyr)
```

### Global Lithium Production

::: {.justify}
The Global Lithium Production dataset was downloaded from the Our World in Data website as a csv file. Very few steps were applied to this dataset due to the quality of the original data.
:::

``` {r}
#| echo: false
#| message: false
#| warning: false

# Save dataframe as a new file
df_production <- read.csv('../../data/01-modified-data/clean_lithium-production.csv')

knitr::kable(head(df_production, 5))
```

[View dataset](https://github.com/anly501/dsan-5000-project-mazuckerberg-git/blob/main/data/01-modified-data/clean_lithium-production.csv)

> [Data Cleaning - Global Lithium Production](https://mazuckerberg.georgetown.domains/DSAN_5000/3_data_cleaning/3_1_lithium_production.html)


### Lithium Companies Stocks

::: {.justify}

Add description
:::

``` {r}
#| echo: false
#| message: false
#| warning: false

# Save dataframe as a new file
df_companies <- read.csv('../../data/01-modified-data/clean_lithium-companies.csv')

knitr::kable(tail(df_companies, 5))
```

[View dataset](https://github.com/anly501/dsan-5000-project-mazuckerberg-git/blob/main/data/01-modified-data/clean_lithium-companies.csv)

> [Data Cleaning - Lithium Companies Stocks](https://mazuckerberg.georgetown.domains/DSAN_5000/3_data_cleaning/3_2_lithium_companies.html)


### Resource Price

::: {.justify}

Add description
:::

``` {r}
#| echo: false
#| message: false
#| warning: false

# Save dataframe as a new file
df_resources <- read.csv('../../data/01-modified-data/clean_resources-price.csv')

knitr::kable(head(df_resources, 5))
```

[View dataset](https://github.com/anly501/dsan-5000-project-mazuckerberg-git/blob/main/data/01-modified-data/clean_resources-price.csv)

> [Data Cleaning - Commodity Price](https://mazuckerberg.georgetown.domains/DSAN_5000/3_data_cleaning/3_4_commodity_price.html)

### Lithium News Sentiment Analysis


::: {.justify}

Add description
:::

``` {r}
#| echo: false
#| message: false
#| warning: false

# Save dataframe as a new file
df_sentiment <- read.csv('../../data/01-modified-data/clean_sentiment_analysis.csv')

knitr::kable(head(df_sentiment, 5))
```

[View dataset](https://github.com/anly501/dsan-5000-project-mazuckerberg-git/blob/main/data/01-modified-data/clean_sentiment_analysis.csv)

> [Data Cleaning - Lithium News Sentiment Analysis](https://mazuckerberg.georgetown.domains/DSAN_5000/3_data_cleaning/3_5_news_api.html)