---
title: "Commodity Price"
format: html
---
```{css, echo = FALSE}
.justify {
    text-align: justify !important;
    text-indent: 20px; 
}

.epigrafe {
    text-align: justify !important;
    text-indent: 20px; 
    border: 1.5px solid #87c8b5; 
    padding-top: 15px;
    padding-bottom: 5px;
    padding-right: 15px;
    padding-left: 15px;
    font-size: 14px;
    background-color: #f9f9f9; 
    margin: 20px 0px 30px 0px;
}
```

``` {r}
#| echo: false
#| message: false
#| code-fold: true
#| code-summary: "Libraries"
#| results: 'hide'
#| warning: false

library(tidyverse)
library(ggplot2)
library(forecast)
library(astsa)
library(xts)
library(tseries)
library(fpp2)
library(fma)
library(lubridate)
library(tidyverse)
library(TSstudio)
library(quantmod)
library(tidyquant)
library(plotly)
library(ggplot2)
library(gridExtra)
library(readxl)
library(zoo)
```

### Uranium Price

::: {.epigrafe}
Before cleaning the dataset, the dataset had ### rows and # columns:

- ###

- ###

- ###

- ###
:::

``` {r}
#| echo: true
#| message: false
#| code-fold: true
#| code-summary: "Data Gathering"
#| warning: false

# Read csv file
df_commodity_price <- read_excel("../../../data/00-raw-data/commodity_price.xlsx")

# Original dataset
head(df_commodity_price, n = 10)
```

**Dimensions:**

``` {r}
#| echo: false
#| message: false
#| warning: false

# Dataset dimensions
dim(df_commodity_price)
```

::: {.epigrafe}
Data Cleaning Description. 
:::

``` {r}
#| echo: true
#| message: false
#| code-fold: true
#| code-summary: "Data Cleaning Code"
#| warning: false

df_commodity_price <- read_excel("../../../data/00-raw-data/commodity_price.xlsx")

df_commodity_price <- df_commodity_price %>%
  pivot_longer(cols = -c('...1'), 
               names_to = "Month_Year",
               values_to = "Price")

df_commodity_price <- df_commodity_price %>% filter(!is.na(Price) & Price != "")

df_commodity_price$Month_Year <- as.yearmon(df_commodity_price$Month_Year, format = "%b %Y")

df_commodity_price$Month_Year <- format(df_commodity_price$Month_Year, "%m-%Y")

df_commodity_price$Month_Year <- paste("01-", df_commodity_price$Month_Year, sep = "")

df_commodity_price$Month_Year <- as.Date(df_commodity_price$Month_Year, format = "%d-%m-%Y")

names(df_commodity_price) <- c('Commodity', 'DATE', 'Price')

df_uranium_price <- df_commodity_price %>%
  filter(Commodity == "Uranium")
```

**Cleaned Data:**

``` {r}
#| echo: false
#| message: false
#| warning: false

# Cleaned dataset
head(df_uranium_price, n = 10)
```

**Summary:**

``` {r}
#| echo: false
#| message: false
#| warning: false

summary(df_uranium_price)
```

**Dimensions:**

``` {r}
#| echo: false
#| message: false
#| warning: false

dim(df_uranium_price)
```

``` {r}
#| echo: false
#| message: false
#| warning: false

# Save dataframe as a new file
write.csv(df_uranium_price, '../../../data/01-modified-data/clean_uranium-price.csv', row.names = FALSE)
```

### Natural Gas Price

::: {.epigrafe}
Before cleaning the dataset, the dataset had ### rows and ### columns:

- ###

- ###

:::

``` {r}
#| echo: true
#| message: false
#| code-fold: true
#| code-summary: "Data Cleaning Code"
#| warning: false

# Set the start and end dates
start_date <- "2000-01-01"
end_date <- "2022-12-31"

# Define the symbol for Gas Price to USD exchange rate
symbol <- "GASREGCOVW"

# Use getSymbols() to fetch the data
getSymbols(symbol, from = start_date, to = end_date, src = "FRED")

# Access the data as a data frame
df_gas_price <- as.data.frame(GASREGCOVW)

df_gas_price <- rownames_to_column(df_gas_price, var = "DATE")

df_gas_price$DATE <- as.Date(df_gas_price$DATE)

# Print the first few rows of the data
head(df_gas_price, n = 10)

```

``` {r}
#| echo: false
#| message: false
#| warning: false

summary(df_gas_price)
```

**Dimensions:**

``` {r}
#| echo: false
#| message: false
#| warning: false

# Dataset dimensions
dim(df_gas_price)
```

::: {.epigrafe}
Data Cleaning Description. 
:::

``` {r}
#| echo: true
#| message: false
#| code-fold: true
#| code-summary: "Data Cleaning"
#| warning: false

# Separate the Date by terms
df_gas_price <- df_gas_price %>%
    mutate(Year = year(DATE)) %>%
    mutate(Month = month(DATE)) %>%
    mutate(Day = day(DATE))


df_gas_price <- df_gas_price %>%
    group_by(Year, Month) %>%
    filter(Day == min(Day)) %>%
    ungroup()


df_gas_price <- df_gas_price %>%
    select(GASREGCOVW, Year, Month, Day)


df_gas_price <- df_gas_price %>%
    mutate(DATE = paste(Year, Month, '01', sep = "-"))


df_gas_price$DATE <- as.Date(df_gas_price$DATE)


df_gas_price <- df_gas_price %>%
    select(GASREGCOVW, DATE)
```

**Cleaned Data:**

``` {r}
#| echo: false
#| message: false
#| warning: false

# Cleaned dataset
head(df_gas_price, n = 10)
```

**Summary:**

``` {r}
#| echo: false
#| message: false
#| warning: false

summary(df_gas_price)
```

**Dimensions:**

``` {r}
#| echo: false
#| message: false
#| warning: false

dim(df_gas_price)
```

``` {r}
#| echo: false
#| message: false
#| warning: false

# Save dataframe as a new file
write.csv(df_gas_price, '../../../data/01-modified-data/clean_gas-price.csv', row.names = FALSE)
```

### Resources Price

::: {.epigrafe}
Data Cleaning Description. 
:::

``` {r}
#| echo: true
#| message: false
#| code-fold: true
#| code-summary: "Data Cleaning"
#| warning: false


df_uranium_price <- df_uranium_price %>% select(DATE, Price)

names(df_uranium_price) <- c('DATE', 'Uranium')

names(df_gas_price) <- c('Natural Gas', 'DATE')

df_resource_price <- merge(df_uranium_price, df_gas_price, by.x = 'DATE', by.y = 'DATE', all = TRUE)
```

**Cleaned Data:**

``` {r}
#| echo: false
#| message: false
#| warning: false

# Cleaned dataset
head(df_resource_price, n = 10)
```

**Summary:**

``` {r}
#| echo: false
#| message: false
#| warning: false

summary(df_resource_price)
```

**Dimensions:**

``` {r}
#| echo: false
#| message: false
#| warning: false

dim(df_resource_price)
```

``` {r}
#| echo: false
#| message: false
#| warning: false

# Save dataframe as a new file
write.csv(df_resource_price, '../../../data/01-modified-data/clean_resources-price.csv', row.names = FALSE)
```